A study on the usage of the wireless signal spectrum in a City of Things network

### 2017-05-25 advnetwpaper/paper_no_ieeetrans.tex
Some changes, generating 'clean' graphs.

@@

@@

% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

% for over three affiliations, or if they all won't fit within the width
% of the page, use this alternative format:
%
%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
%Homer Simpson\IEEEauthorrefmark{2},
%James Kirk\IEEEauthorrefmark{3},
%Montgomery Scott\IEEEauthorrefmark{3} and
%Eldon Tyrell\IEEEauthorrefmark{4}}
%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
%Georgia Institute of Technology,
%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
%Email: homer@thesimpsons.com}
%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212, Fax: (888) 555--1212}
%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
The abstract goes here.
\end{abstract}

\section{Introduction}
% no \IEEEPARstart
Most of the radio spectrum requires a license to operate on. Devices emitting Radio Frequency (RF) energy without a license are limited to using one of the unlicensed radio bands, known as the Industrial, Scientific and Medical (ISM) radio bands. The most well-known of these bands is the 100MHz wide band centered around 2.45GHz, often called the 2.4GHz band. Many consumer devices such as wireless mice and keyboards, bluetooth headphones and Radio Controlled (RC) cars communicate in the 2.4GHz band. Furthermore most IEEE 802.11 (Wi-Fi) routers operate in the 2.4GHz band, along with the 5GHz band (which actually ranges from 5.725GHz to 5.875GHz). These ISM bands were originally intended and still used for devices emitting RF energy for purposes other than communication. The most common such application is the microwave oven. All this leads to high interference in these bands. In this paper, we investigate how bad this 2.4GHz and 5GHz interference actually is in the city of Antwerp. \\ \\
To measure this interference we use the City of Things (CoT) network, operated by research institute imec. This network consists of hundreds of sensors and wireless gateways positioned around the city of Antwerp. On 13 of these gateways, we sampled activity on the 2.4GHz and 5GHz bands every 10 minutes for TODO days. The main motivation of this paper was in fact to investigate whether interference around the gateways is significant enough to pose a problem for the network. Next to the general magnitude of the interference, we look for patterns in daily and weekly variation of the interference.

Some citing \cite{adams4242save} % At least one citation is needed, otherwise IEEEtran gives a nice error.

\section{Measuring Interference}
All used nodes have a COMPEX WLE900VX-7A network adapter, containing a Qualcomm Atheros QCA9880 wireless chipset. This 802.11ac chipset, along with all other 802.11ac and 802.11n chips by Qualcomm Atheros support a mode called \textit{spectral scan}. In this mode the chip can scan the activity on all frequencies it supports. This activity is not limited to 802.11 traffic, but is influenced by any received signal. In this mode, the chip acts as a simple spectrum analyzer. The Free and Open-Source Software (FOSS) ath9k and ath10k drivers, respectively for the 802.11n and 802.11ac Qualcomm Atheros chipsets, support this mode. Each scan is performed over a period of \SI{4}{\micro\second}. Every scanned channel is divided into 64, 128 or 256 equally wide bins. For the full channel the noise floor and Received Signal Strength Indicator (RSSI) are reported, along with a magnitude for each bin. This magnitude incidates  how the power within the channel was divided across the bins. We provide an overview of how to interpret these values in TODO. \\
\subsection{Settings}
The ath10k version of spectral scan offers fewer options than the ath9k version. For instance the option to increase the scan time from \SI{4}{\micro\second} to \SI{2044}{\micro\second} is absent in ath10k. The only missing option relevant to our case is the \textit{chanscan} mode of spectral scan. When this mode is triggered the driver gathers a configurable number of samples for every channel it supports. We emulate this behaviour using the \textit{background} mode, which continuously scans the chipset's currently configured channel as long as it is not sending or receiving. We combine this with an \textit{iw scan} command, which listens for access point beacons for a provided list of frequencies. TODO PASSIVE When providing \textit{iw scan} with a list of all supported frequencies, this closely emulates the \textit{chanscan} behaviour.

@@

Before going into the interpretation of this specific data, we provide a quick overview of the dBm unit, often used to express power ratios. The regular decibel (dB) unit is a dimensionless unit. In contrast dBm expresses absolute power in reference to watt. 0dBm is defined as 1 milliwatt (mW). For each increase of 3dBm, the power in mW doubles, and for each increase of 10dBm, the power is multiplied by 10. Milliwatt-to-dBm conversion is calculated using the formula $10 log_{10}(\frac{x}{\SI{1}{\milli\watt}})$ with $x$ the power in mW. A negative dBm power simply means the power is less than \SI{1}{\milli\watt}. By using dBm, very small or large power values are often avoided. As power attenuates quadratically with distance, useful power values can vary dramatically.\\\\
For each channel scan, the driver reports the noise floor and RSSI value. The noise floor value is hard-coded for every channel and ranges between -95dBm and -106dBm. This value is the expected power of all noise in a channel combined. The chipset can only receive a singal whose power exceeds this noise floor. The RSSI value is a (unitless) integer indicating, as the name implies, the received power in the channel. While the existence of this value is part of the 802.11 specification, its calculation is not. As a result every manufacturer calculates the RSSI differently. Atheros chips follow a rather simple formula: subtracting the noise floor from the current signal strength in dBm results in the RSSI. These two values are enough to estimate the signal strength in a channel: a noise floor of -96dBm and an RSSI of 40 indicate a received signal of -56dBm. \\

@@

As the dBm power values computed from the noise floor and RSSI are not linear but rather logarithmic, we cannot simply multiply by $c(i)$ to get each bin's power value. Instead we first convert $c(i)$ to a logarithmic value: $c_{log}(i) = 10 log_{10}(c(i))$. As $log(a*b) = log(a) + log(b)$, we add this coefficient to the total logarithmic power to calculate each share's power: $nf + RSSI + c_{log}(i)$. Noting that $log(a/b) = log(a) - log(b)$, writing the formula using only the parameters returned by the spectral scan results in $nf + RSSI  + 10 log_{10}(b(i)^2) - 10 log_{10}(\sum\limits_{j=1}^{n}{b(j)^2})$.

@@

\subsection{Subsection Heading Here}
Subsection text here.

\subsubsection{Subsubsection Heading Here}
Subsubsection text here.

\section{Measurements}
At first we measured the spectrum on every channel in both the 2.4GHz and the 5GHz band every 10 minutes. Looking at interpretations of the resulting data it was clear that a shorter time between scans would be useful because recurring spikes were missed on some days while they definitely there and also quite interesting.
\\< images waarop da duidelijk is >\\
We then reduced the time between scans from 10 minutes to 2 minutes, which was the minimum we could achieve while guaranteeing that scans would be succesful.
Using an adapted version of Simon Wünderlich's fft_eval code, GNUplot and Savitzky-Golay (source? TODO) filtering we graphically represent our data points per node.
<tussen maandag 15 tem zondag 21 mei>

\section{Findings}

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex,
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation results for the network.}
%\label{fig_sim}
%\end{figure}

% Note that the IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command,
% and the \label for the overall figure must come after \caption.
% \hfil is used as a separator to get equal spacing.
% Watch out that the combined width of all the subfigures on a
% line do not exceed the text width or a line break will occur.
%
%\begin{figure*}[!t]
%\centering
%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
%\label{fig_second_case}}
%\caption{Simulation results for the network.}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat[]), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.
% Be aware that for subfig.sty to generate the (a), (b), etc., subfigure
% labels, the optional argument to \subfloat must be present. If a
% subcaption is not desired, just leave its contents blank,
% e.g., \subfloat[].

% An example of a floating table. Note that, for IEEE style tables, the
% \caption command should come BEFORE the table and, given that table
% captions serve much like titles, are usually capitalized except for words
% such as a, an, and, as, at, but, by, for, in, nor, of, on, or, the, to
% and up, which are usually not capitalized unless they are the first or
% last word of the caption. Table text will default to \footnotesize as
% the IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that the IEEE does not put floats in the very first column
% - or typically anywhere on the first page for that matter. Also,
% in-text middle ("here") positioning is typically not used, but it
% is allowed and encouraged for Computer Society conferences (but
% not Computer Society journals). Most IEEE journals/conferences use
% top floats exclusively.
% Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the
% \fnbelowfloat command of the stfloats package.

\section{Conclusion}
The conclusion goes here.

% conference papers do not normally have an appendix

% use section* for acknowledgment
\section*{Acknowledgment}

The authors would like to thank...

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,bib}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)

>>>>

%- algemene verloop doorheen dag/week
%- recurring patterns tussen dagen
%- drukkere/rustigere channels
%- 2.4 vs 5 (gelijkaardig aan het vorige)

@@

\usepackage{hyperref}

@@

How bad is it, in general? 
Are there patterns, e.g. daily recurrence? 
Advanced: how can you ensure data quality, i.e., how to validate the results?

\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract

\tableofcontents
\clearpage

\section{Introduction}
% no \IEEEPARstart
Most of the radio spectrum requires a license to operate on. Devices emitting Radio Frequency (RF) energy without a license are limited to using one of the unlicensed radio bands, known as the Industrial, Scientific and Medical (ISM) radio bands. The most well-known of these bands is the 100MHz wide band centered around 2.45GHz, often called the 2.4GHz band. Many consumer devices such as wireless mice and keyboards, bluetooth headphones and Radio Controlled (RC) cars communicate in the 2.4GHz band. Furthermore most IEEE 802.11 (Wi-Fi) routers operate in the 2.4GHz band, along with the 5GHz band (which actually ranges from 5.725GHz to 5.875GHz). These ISM bands were originally intended and still used for devices emitting RF energy for purposes other than communication. The most common such application is the microwave oven. All this leads to high interference in these bands. In this paper, we investigate how bad this 2.4GHz and 5GHz interference actually is in the city of Antwerp. \\ \\
To measure this interference we use the City of Things (CoT) network, operated by research institute imec. This network consists of hundreds of sensors and wireless gateways positioned around the city of Antwerp. On 11 of these gateways, we sampled activity on the 2.4GHz and 5GHz bands every 2 minutes for a little over 10 days, to ensure that we would have at least one full week of usable data for each node. The main motivation of this paper was in fact to investigate whether interference around the gateways is significant enough to pose a problem for the network. Next to the general magnitude of the interference, we look for patterns in daily and weekly variation of the interference.
\section{Measuring Interference}
All used nodes have a COMPEX WLE900VX-7A network adapter, containing a Qualcomm Atheros QCA9880 wireless chipset. This 802.11ac chipset, along with all other 802.11ac and 802.11n chips by Qualcomm Atheros support a mode called \textit{spectral scan}. In this mode the chip can scan the activity on all frequencies it supports. This activity is not limited to 802.11 traffic, but is influenced by any received signal. In this mode, the chip acts as a simple spectrum analyzer. The Free and Open-Source Software (FOSS) ath9k and ath10k drivers, respectively for the 802.11n and 802.11ac Qualcomm Atheros chipsets, support this mode. Each scan is performed over a period of \SI{4}{\micro\second}. Every scanned channel is divided into 64, 128 or 256 equally wide bins. For the full channel the noise floor and Received Signal Strength Indicator (RSSI) are reported, along with a magnitude for each bin. This magnitude incidates  how the power within the channel was divided across the bins. We provide an overview of how to interpret these values in \mbox{\hyperref[sec:interpret]{Interpreting spectral scans}}. \\
\subsection{Settings}
The ath10k version of spectral scan offers fewer options than the ath9k version. For instance the option to increase the scan time from \SI{4}{\micro\second} to \SI{2044}{\micro\second} is absent in ath10k. The only missing option relevant to our case is the \textit{chanscan} mode of spectral scan. When this mode is triggered the driver gathers a configurable number of samples for every channel it supports. We emulate this behaviour using the \textit{background} mode, which continuously scans the chipset's currently configured channel as long as it is not sending or receiving. We combine this with an \textit{iw scan} command, which listens for access point beacons for a provided list of frequencies. When providing \textit{iw scan} with a list of all supported frequencies, this closely emulates the \textit{chanscan} behaviour.
\label{sec:interpret}

@@

Before going into the interpretation of this specific data, we provide a quick overview of the dBm unit, often used to express power ratios. The regular decibel (dB) unit is a dimensionless unit. In contrast dBm expresses absolute power in reference to watt. 0 dBm is defined as 1 milliwatt (mW). For each increase of 3dBm, the power in mW doubles, and for each increase of 10dBm, the power is multiplied by 10. Milliwatt-to-dBm conversion is calculated using the formula\\ \\ $dBm = 10 log_{10}(\dfrac{x}{\SI{1}{\milli\watt}})$ with $x$ the power in mW.\\ \\ A negative dBm power simply means the power is less than \SI{1}{\milli\watt}. By using dBm, very small or large power values are often avoided. As power attenuates quadratically with distance, useful power values can vary dramatically.\\\\
For each channel scan, the driver reports the noise floor and RSSI value. The noise floor value is hard-coded for every channel and ranges between -95dBm and -106dBm. This value is the expected power of all noise in a channel combined. The chipset can only receive a singal whose power exceeds this noise floor. The RSSI value is a (unitless) integer indicating, as the name implies, the received power in the channel. While the existence of this value is part of the 802.11 specification, its calculation is not. As a result every manufacturer calculates the RSSI differently. Atheros chips follow a rather simple formula: subtracting the noise floor from the current signal strength in dBm results in the RSSI. These two values are enough to estimate the signal strength in a channel: a noise floor of -96dBm and an RSSI of 40 indicate a received signal of -56dBm. TODO SOURCE ? (for example :http://www.n-cg.net/ncgpdf/WiFi\_SignalValues.pdf see conversion for atheros) \\

@@

As the dBm power values computed from the noise floor and RSSI are not linear but rather logarithmic, we cannot simply multiply by $c(i)$ to get each bin's power value. Instead we first convert $c(i)$ to a logarithmic value: \\ \\ $c_{log}(i) = 10 log_{10}(c(i))$. \\ \\ As $log(a*b) = log(a) + log(b)$, we add this coefficient to the total logarithmic power to calculate each share's power:\\ \\ $nf + RSSI + c_{log}(i)$.\\ \\ Noting that $log(a/b) = log(a) - log(b)$, writing the formula using only the parameters returned by the spectral scan results in\\ \\ $nf + RSSI  + 10 log_{10}(b(i)^2) - 10 log_{10}(\sum\limits_{j=1}^{n}{b(j)^2})$ \\ \\

@@

\section{Measurements}
At first we measured the spectrum on every channel in both the 2.4GHz and the 5GHz band every 10 minutes. Looking at graphic interpretations of the resulting data it was clear that a shorter time between scans would be more effective at showing us the state of the spectrum because recurring spikes were missed on some days despite their presence being clear in subsequent smaller tests. 
\\< images waarop da duidelijk is >\\
Another point that was raised following the first set of graphic interpretations was that of outlier detection and elimination, occasionally there were big spikes in the received signal strength which are most likely caused by interference.
Playing around with some parameters, we reduced the time between scans from 10 minutes to 2 minutes and discard the 100 largest readings for every sample (when looking at all the channels). This limit of 2 minutes was the minimum we could achieve while guaranteeing a scan would succeed and the number of samples to discard was chosen based on some experimentation.
Using an adapted version of Simon Wünderlich's fft$\_$eval code, Python's Matplotlib library and Savitzky-Golay (source? TODO) filtering we graphically represent our data points per node. For CoT-node12 (location : UAntwerpen campus Middelheim, indoors), looking at the 2.4 GHz band on every channel from monday May 15th to monday May 22nd, we got the following graph:\\
\includegraphics[scale=0.7]{images/window_361_2.4/2017-05-22cot-node12-student_image.pdf}\\
Each line/color represents a day, using this representation we tried to find recurring patterns in the measurements for every node separately. When looking at the complete set of channels, patterns were not as clear as we'd hoped.\\
Given this and the fact that a large amount of static on one channel would make it seem like the entire spectrum is full we switched to drawing the graphs per non-overlapping channel and observing the results in that way. This lead to following graph on which we can clearly see that on weekdays there is substantially more traffic during the daytime than on the weekend (which is as expected for a university campus):\\
\includegraphics[scale=0.7]{images/2017-05-22cot-node12-student_image.pdf}\\

\section{Findings}
\subsection{Evolution throughout the day}
\subsection{Evolution throughout the week}
\subsection{Recurring patterns between days} 
On the different nodes we had varying degrees of success in finding recurring patterns. For this report we've only included the nodes on which we did find such patterns (and we have included the non pattern-y graphs in the zip file that is included // they are available on demand? TODO) 
\subsection{Calm and busy channels}
1 busy, 6, 11 calm
\subsection{2.4 GHz vs 5 GHz}
\subsection{Validation}
scans met / zonder sturende laptop - stukje had gij al
dagelijkse patronen
outside / inside t zelfde +-

\section{Conclusions}
The conclusion goes here.
